Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data

### 2022-08-18 analysis/paper/paper.Rmd
abstract done

---
title: "Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data"
author:
  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^{a,b}$, ...
  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Potsdam, Germany*
  - $^b$ *Deutsches Archäologisches Institut, Dendroclimatology, Berlin, Germany*
  - "* Corresponding author: hurley@gfz-potsdam.de"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::html_document2:
      fig_caption: yes
      toc: true
      number_sections: false
    bookdown::word_document2:
      number_sections: true
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      toc: yes
      number_sections: true
header-includes:
- \usepackage{amsmath}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
bibliography: 
  - uhi_references.bib
  - 04_gfz_UHI.bib
  - 07_phenoTram.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The socio-ecological benefits that urban trees provide depend on their health and performance. Excess heat (i.e., Urban Heat Island; UHI) and other urban conditions affect tree physiology with outcomes from enhanced growth to mortality. Resilient urban forests in the face of climate change require species-specific understanding of growth responses. However, to date, studies assessing growth dynamics were primarily based on remote sensing of communities rather than individuals, or relied on labor-intensive methods that can limit the spatial coverage necessary to account for highly variable urban growing conditions. Here, we analyze growth dynamics of common urban tree species over time and across space for Berlin (Germany), a large metropolis, combining dendroecological (temporal) and inventory assessments (spatial). First, we show that annual increments increased across the 20th century for early (i.e., young) growth. Second, we use an approach relying on open inventory data to identify growth potential (diameter) in relation to excess heat across Berlin while accounting for age, potential management effects, and the urban fabric (e.g., planting area; building density, height; available soil nutrients) by applying generalized additive models for the ten most abundant species. Our analyses showed that younger trees may benefit from increased temperatures, while older individuals feature lower growth at greater UHI magnitudes. Furthermore, we show that planting area as well as building density modulate growth responses to temperature. Lastly, we discuss implications for the management of urban trees in the context of climate change mitigation, considering that younger trees are predominantly located at UHI “hot spots” and will undergo the observed age-dependent shift in temperature-growth sensitivity. By relying on increasingly available open data (inventories, urban/environmental data) our approach here is or will be readily transferable to other urban regions, allowing for species-specific growth assessments, while accounting for urban conditions at an individual level.

@@

### Temperature/UHI data

Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview)).
Two urban air temperature and one surface UHI data set were tested as explanatory variables in generalized additive models (see Section \@ref(sec:stat-gam)).
The air temperatures from the Berlin environmental atlas (EnvAt) are processed model outputs [FITNAH-3D; cf. @gross1994] that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted after rasterizing ($5~m$ resolution).
UrbClim air temperatures are hourly model outputs [100$~m$ resolution, @deridder2015] based on ERA5 re-analyses data (ECMWF) for which observations from the hottest month  available (June, 2011) were averaged to hours equivalent to Berlin Environmental Atlas (referred to as Berlin EnvAt) data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
Subsequently, a land-use and land-cover mask (CORINE; European Union, Copernicus Land Monitoring Service 2018, European Environment Agency) was used to define urban and rural/forested areas.
Using this mask was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
The urban heat loading was then calculated as

@@

<!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
From all tested models the most suitable (i.e., parsimonious with highest explanatory power) was employed for further analyses.

@@

where $g()$ is a log-link for $DBH \sim Gamma$, and $i$, $j$ are indices for observations and species, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or LandSat at different times (morning, afternoon/day, night; cf. Section$~$\@ref(sec:dsources)); $c$ is a species-dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ districts to account for differing management regimes across the city.
A global spatial smooth $f(x_i,y_i)$ (representing projected coordinates in UTM) via a Gaussian process [cf. p. 242 in @wood2017] was included to account for the spatial structure of observations, which reduced auto-correlation of model residuals considerably (see supplemental information).
These were compared to a sub-set of models without a spatial component (cf. Equation$~$\@ref(eq:gam-nocovariate), and Figure$~$\@ref(fig:fig-model-deviance) but not further discussed there).
We also tested a suite of models without the spatial smooth for comparison.

@@

(ref:cambial-effect) Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section$~$\@ref(sec:methods-dendro)). Thick lines and bands are for mean and 95$~\%$ confidence interval of all annual predictions across a time group (fine lines). On average, early stages of recent growth (up to approximately 22 years) exceed that of earlier periods discernibly.

@@

(ref:lcz6-effect) Age-group averaged relationships between the LCZ6 and stem diameter for two $Tilia$ genera from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LandSat-derived UHI as excess heat measure). LCZ6 cover was derived as a weighted average within a radius of 150$~m$ around each tree. Note the identical shape, but difference in intercept as a result of the model specification with a global smooth the non-temperature covariate for all species (see Section$~$\@ref(sec:methods-gam-heat)); the UHI magnitude was fixed at $3^\circ C$ where most trees are present across age groups. The model identified a steady increase of average stem diameter with open to mid-rise urban land cover.

@@

<!-- <!-- The following line inserts a page break  --> 
<!-- `r pgbreak()` -->

>>>>

---
title: "Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data"
author:
  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^{a,b}$, ...
  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Potsdam, Germany*
  - $^b$ *Deutsches Archäologisches Institut, Dendroclimatology, Berlin, Germany*
  - "* Corresponding author: hurley@gfz-potsdam.de"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::html_document2:
      fig_caption: yes
      toc: false
      number_sections: false
      global_numbering: true
    bookdown::word_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
      global_numbering: true
    bookdown::pdf_document2:
      fig_caption: yes
      toc: yes
      number_sections: false
      global_numbering: true
header-includes:
- \usepackage{amsmath}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
bibliography: 
  - uhi_references.bib
  - 04_gfz_UHI.bib
  - 07_phenoTram.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The socio-ecological benefits that urban trees provide depend on their health and performance. Excess heat (i.e., Urban Heat Island; UHI) and other urban conditions affect tree physiology with outcomes from enhanced growth to mortality. Resilient urban forests in the face of climate change require species-specific understanding of growth responses. However, to date, studies assessing growth dynamics were primarily based on remote sensing of communities rather than individuals, or relied on labor-intensive methods that can limit the spatial coverage necessary to account for highly variable urban growing conditions. Here, we analyze growth dynamics of common urban tree species over time and across space for Berlin (Germany), a large metropolis, combining dendroecological (temporal) and inventory assessments (spatial). First, we show that annual increments increased across the 20th century for early (i.e., young) growth. Second, we use an approach relying on open inventory data to identify growth potential (diameter) in relation to excess heat across Berlin while accounting for age, potential management effects, and the urban fabric (e.g., planting area; building density, height; available soil nutrients) by applying generalized additive models for the ten most abundant species. Our analyses showed that younger trees may benefit from increased temperatures, while older individuals feature lower growth at greater UHI magnitudes. Furthermore, we show that planting area as well as building density modulate growth responses to temperature. Lastly, we discuss implications for the management of urban trees in the context of climate change mitigation, considering that younger trees are predominantly located at UHI “hot spots” and will undergo the observed age-dependent shift in temperature-growth sensitivity. By relying on increasingly available open data (inventories, urban/environmental data) our approach here is or will be readily transferable to other urban regions, allowing for species-specific growth assessments, while accounting for urban conditions at an individual level.

@@

### Temperature/UHI data {#sec:tempdata}

Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview)).
Two urban air temperature and one surface UHI data set were tested as explanatory variables in generalized additive models (see Section [GAMs](#sec:stat-gam)).
The air temperatures from the Berlin environmental atlas (EnvAt) are processed model outputs [FITNAH-3D; cf. @gross1994] that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted after rasterizing ($5~m$ resolution).
UrbClim air temperatures are hourly model outputs [100$~m$ resolution, @deridder2015] based on ERA5 re-analyses data (ECMWF) for which observations from the hottest month  available (June, 2011) were averaged to hours equivalent to Berlin Environmental Atlas (referred to as Berlin EnvAt) data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
Subsequently, a land-use and land-cover mask (CORINE; European Union, Copernicus Land Monitoring Service 2018, European Environment Agency) was used to define urban and rural/forested areas.
Using this mask was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
The urban heat loading was then calculated as

@@

<!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section [GAMs](#sec:stat-gam) for details).
Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
From all tested models the most suitable (i.e., parsimonious with highest explanatory power) was employed for further analyses.

@@

where $g()$ is a log-link for $DBH \sim Gamma$, and $i$, $j$ are indices for observations and species, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or LandSat at different times (morning, afternoon/day, night; see Section [Temperature/UHI data](#sec:tempdata); $c$ is a species-dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ districts to account for differing management regimes across the city.
A global spatial smooth $f(x_i,y_i)$ (representing projected coordinates in UTM) via a Gaussian process [cf. p. 242 in @wood2017] was included to account for the spatial structure of observations, which reduced auto-correlation of model residuals considerably (see supplemental information).
These were compared to a sub-set of models without a spatial component (cf. Equation$~$\@ref(eq:gam-nocovariate), and Figure$~$\@ref(fig:fig-model-deviance) but not further discussed there).
We also tested a suite of models without the spatial smooth for comparison.

@@

(ref:cambial-effect) Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section [Dendrochronological analyses](#sec:methods-dendro). Thick lines and bands are for mean and 95$~\%$ confidence interval of all annual predictions across a time group (fine lines). On average, early stages of recent growth (up to approximately 22 years) exceed that of earlier periods discernibly.

@@

(ref:lcz6-effect) Age-group averaged relationships between the LCZ6 and stem diameter for two $Tilia$ genera from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LandSat-derived UHI as excess heat measure). LCZ6 cover was derived as a weighted average within a radius of 150$~m$ around each tree. Note the identical shape, but difference in intercept as a result of the model specification with a global smooth the non-temperature covariate for all species (see Section [Stem diameter model development and selection](#sec:methods-gam-heat); the UHI magnitude was fixed at $3^\circ C$ where most trees are present across age groups. The model identified a steady increase of average stem diameter with open to mid-rise urban land cover.

@@

# Appendices
Information too detailed to be included in the main text, for instance a list of areas sampled, may be presented in appendices which are to be placed before the References. Appendices should be as short and succinct as possible.

<!-- <!-- The following line inserts a page break  --> 
`r pgbreak()`
